---
title: "Hepatitis B (HepB3) and Hib3 Vaccination Coverage in 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 5)

local_lib <- file.path(getwd(), "r_libs")
if (dir.exists(local_lib)) {
  .libPaths(c(local_lib, .libPaths()))
}

library(tidyverse)
library(ggplot2)
library(maps)

name_map <- c(
  "UAE" = "United Arab Emirates",
  "Cote d'Ivoire" = "Ivory Coast",
  "Congo, Dem. Rep." = "Democratic Republic of the Congo",
  "Congo, Rep." = "Republic of Congo",
  "Micronesia, Fed. Sts." = "Micronesia",
  "Kyrgyz Republic" = "Kyrgyzstan",
  "St. Kitts and Nevis" = "Saint Kitts",
  "Lao" = "Laos",
  "St. Lucia" = "Saint Lucia",
  "Slovak Republic" = "Slovakia",
  "Eswatini" = "Swaziland",
  "St. Vincent and the Grenadines" = "Saint Vincent"
)

read_vacc <- function(path, indicator_label) {
  read.csv(path, check.names = FALSE) %>%
    rename(geo = geo, country = name) %>%
    pivot_longer(
      cols = -c(geo, country),
      names_to = "year",
      values_to = "coverage"
    ) %>%
    mutate(
      year = as.integer(year),
      coverage = as.numeric(coverage),
      indicator = indicator_label
    )
}

plot_world_map <- function(data_2020, title, fill_label) {
  map_join_name <- if_else(
    data_2020$country %in% names(name_map),
    name_map[data_2020$country],
    data_2020$country
  )

  plot_data <- data_2020 %>%
    mutate(map_region = map_join_name) %>%
    right_join(map_data("world"), by = c("map_region" = "region"))

  ggplot(plot_data, aes(x = long, y = lat, group = group, fill = coverage)) +
    geom_polygon(color = "grey85", linewidth = 0.1) +
    coord_fixed(ratio = 1.3) +
    scale_fill_viridis_c(
      option = "C",
      na.value = "grey90",
      limits = c(0, 100),
      name = fill_label
    ) +
    theme_void() +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", hjust = 0.5)
    ) +
    labs(title = title, x = NULL, y = NULL)
}
```

## Introduction

This report compares national **Hepatitis B third-dose (HepB3)** and **Haemophilus influenzae type b third-dose (Hib3)** vaccination coverage in **2020**, using country-level estimates from the provided datasets.

## Data overview

```{r load-data}
hepb_raw <- read.csv("hepb3_vacc.csv", check.names = FALSE)
hib_raw <- read.csv("hib3_vacc.csv", check.names = FALSE)

hepb <- read_vacc("hepb3_vacc.csv", "HepB3")
hib <- read_vacc("hib3_vacc.csv", "Hib3")
```

A sample of each source file:

```{r sample-hepb}
knitr::kable(head(hepb_raw, 6))
```

```{r sample-hib}
knitr::kable(head(hib_raw, 6))
```

## 2020 coverage

Each indicator is restricted to **2020**; countries without a reported value are excluded from numeric summaries but still appear as missing on the maps.

```{r subset-2020}
hepb_2020 <- hepb %>% filter(year == 2020)
hib_2020 <- hib %>% filter(year == 2020)

combined_2020 <- hepb_2020 %>%
  select(geo, country, hepb3 = coverage) %>%
  inner_join(
    hib_2020 %>% select(geo, hib3 = coverage),
    by = "geo"
  ) %>%
  filter(!is.na(hepb3), !is.na(hib3)) %>%
  mutate(
    # Per-country estimate of receiving both third doses (independence assumption)
    both_est_pct = (hepb3 / 100) * (hib3 / 100) * 100
  )

summary_2020 <- tibble(
  Indicator = c(
    "HepB3",
    "Hib3",
    "HepB3 (countries with both indicators)",
    "Hib3 (countries with both indicators)",
    "Both vaccines — estimated joint coverage"
  ),
  Countries = c(
    sum(!is.na(hepb_2020$coverage)),
    sum(!is.na(hib_2020$coverage)),
    nrow(combined_2020),
    nrow(combined_2020),
    nrow(combined_2020)
  ),
  Median_coverage = c(
    median(hepb_2020$coverage, na.rm = TRUE),
    median(hib_2020$coverage, na.rm = TRUE),
    median(combined_2020$hepb3),
    median(combined_2020$hib3),
    median(combined_2020$both_est_pct)
  ),
  Mean_coverage = c(
    mean(hepb_2020$coverage, na.rm = TRUE),
    mean(hib_2020$coverage, na.rm = TRUE),
    mean(combined_2020$hepb3),
    mean(combined_2020$hib3),
    mean(combined_2020$both_est_pct)
  )
)

knitr::kable(summary_2020, digits = 1)
```

Among the `r nrow(combined_2020)` countries with 2020 data for **both** indicators, the estimated share of children receiving **both** HepB3 and Hib3 is **`r sprintf("%.1f", mean(combined_2020$both_est_pct))`%** on average (median **`r sprintf("%.1f", median(combined_2020$both_est_pct))`%**). This uses each country’s formula (HepB3% × Hib3% ÷ 100), then averages across countries; it is an unweighted country mean, not a global population total.

## World maps (2020)

### HepB3 vaccination coverage

```{r map-hepb}
plot_world_map(
  hepb_2020 %>% filter(!is.na(coverage)),
  "HepB3 vaccination coverage by country, 2020",
  "Coverage (%)"
)
```

### Hib3 vaccination coverage

```{r map-hib}
plot_world_map(
  hib_2020 %>% filter(!is.na(coverage)),
  "Hib3 vaccination coverage by country, 2020",
  "Coverage (%)"
)
```

## Relationship between HepB3 and Hib3 (2020)

```{r scatter}
cor_2020 <- cor(combined_2020$hepb3, combined_2020$hib3)

ggplot(combined_2020, aes(x = hepb3, y = hib3)) +
  geom_point(alpha = 0.7, color = "#2166AC") +
  geom_smooth(method = "lm", se = TRUE, color = "#B2182B", fill = "#FDDBC7") +
  labs(
    title = "HepB3 vs Hib3 vaccination coverage, 2020",
    x = "HepB3 coverage (%)",
    y = "Hib3 coverage (%)"
  ) +
  coord_equal(xlim = c(0, 100), ylim = c(0, 100)) +
  theme_minimal(base_size = 12)
```

Among `r nrow(combined_2020)` countries with both indicators reported in 2020, the Pearson correlation is **`r round(cor_2020, 2)`**.

## Interpretation

HepB3 and Hib3 coverage in 2020 are **strongly positively associated**: countries with higher HepB3 coverage tend to have higher Hib3 coverage, and the scatterplot points cluster close to the diagonal. The high correlation (`r round(cor_2020, 2)`) suggests that both indicators largely reflect the same underlying strength of routine immunization systems rather than independent program performance.

The world maps show similar geographic patterns for both vaccines—high coverage across much of the Americas, Europe, and parts of Asia, and lower coverage in several countries in sub-Saharan Africa and a few other regions. Large gaps on the maps (grey countries) indicate missing 2020 estimates rather than zero coverage.

Together, the maps and scatterplot indicate that **gains in one third-dose vaccine are typically accompanied by gains in the other**, which is consistent with both being delivered through routine childhood immunization schedules in the same health systems.
